### 2021-04-12 paper.Rmd
wordsmithing

The focus of this review will be on visualizing high-dimensional numerical data using linear projections, in particular, as provided by the grand tour [@Asimov1985-xr; @Buja2005-cx]. The reason being that it is not feasible to adequately review the very large area of visualizing high-dimensions, and there have been numerous developments in tours recently. An overview of the technique, and new modifications is provided, along with how these techniques can be used in a variety of applications, and to help understand nonlinear dimension reduction results.

While tours are invaluable for assessing the geometry of data, they are by no means the only technique available for visualizing structure in high dimensional data. An early technique proposed for assessing pairwise relationships between variables is the scatter plot matrix (SPLOM) [@Tukey1981-ov; @Chambers1983-dh; @Tukey1983-fj; @Carr1984-xw; @Becker1987-dw]. The SPLOM allows the viewer to assess correlation structure but does not scale to large numbers of variables. Similarly, parallel coordinates plots (PCP) can be used to explore correlation and collinearity [@Inselberg1985-kf; @Wegman1990-dy]. By placing multiple variables side by side in a PCP higher order structure like clustering or lower dimensional embeddings are revealed, however the ordering of variables along the axis changes what can be learned. Another display that relies on variable ordering is the heatmap, which is widely used to visualize cluster structure in bioinformatics. @Wilkinson2009-pj provides a comprehensive history of this display in the social and natural sciences.

All of the aforementioned techniques can be enhanced through the use of interactivity.
By combining views with interactive elements like tool tips or highlighting the analysts is able to quickly interrogate interesting features of the data. One  particularly important interaction technique in the history of statistical graphics is called brushing [@Becker1987-dw]. When brushing an analyst drags their mouse over the view which results in a region being drawn onto the canvas. When there are multiple views present the act of brushing can be thought of as a database query; points that fall inside the brush can be used as a look-up query to highlight or filter data on adjacent views (Figure \ref{fig:brushing}). This technique is particularly useful when combined with the tour (Section \@ref(sec:spin)).

@@

Below introduces the notations used for describing $d$-dimensional projections in a $p$-dimensional space. Let $X_{n \times p}$ be the data whose projections are of interest. A projection basis, $A_{p \times d}$, characterizes the direction from which the data are projected and needs to satisfy an orthonormality condition, which requires each column to be a unit vector and to be perpendicular to each other. This condition gives the $p$-dimensional basis space a geometrical structure of a Stiefel manifold. With a data matrix and a projection basis, a projection of the data can be defined as $Y = X \cdot A$. Figure \ref{fig:notation-huber} shows an example of the 2D palmerpenguins data (bill measures only) in a Huber plot, along with a 1D projection in the histogram. This projection is created based on projecting the data onto the solid line in the Huber plot and the it shows the separation among three species.

\begin{figure*}
   \centering
         \includegraphics[width=\textwidth]{figures/notation-huber.pdf}
\caption{Huber's plot (left) of the 2D penguins data (bill length nad depth only). The solid ellipse represents how well the projection, onto different directions, separating the three species and the dashed circle is a reference guide. The histogram (right) is the 1D projection of the data onto the direction outlined as the solid line in the huber's plot. Adapted from E.-K. Lee et al. (2005), Figure 1. }
\label{fig:notation-huber}
\end{figure*}

@@

Various displays are available to show projections in 1D, 2D, or higher dimensions. A 1D projection displays the data in a histogram as if a hand shadow being projected onto a wall. Two-d projections are usually shown as scatterplot since it is the most common display for bivariate relationship. Higher dimensional projections can be shown with multivariate display like scatterplot matrix or parallel coordinates plot. More display methods can be found in [@Wickham2011-uz]. Figure \ref{fig:display} shows a variety of displays for projections and these displays are useful to show the non-normal distributions in the projected data, clustering structure, and multivariate relationship between variables.

@@

@@

Instead of randomly generating targets, guided tour [@Cook1995-ae] finds its targets with a technique called projection pursuit [@Friedman1974-ck].  In projection pursuit, an index function is a surjective mapping from the projection $Y$ in $\mathbb{R}^{n \times d}$ to a scalar in $\mathbb{R}$. Guided tour iteratively optimizes the index function in the projection pursuit and produces the most interesting projection according to the index measure. The right panel of Figure \ref{fig:notation-target-gen} shows an example of a guided tour path.

@@

The smooth animation between the selected target frames is generated via a step-wise rotation in the high-dimensional space. Different methods to derive this rotation were described in @Buja2005-cx, and we can distinguish two types of interpolation [EW: ??]. When interpolating between frames we rotate the starting basis onto the exact basis representation selected as a basis, and @Buja2005-cx describes three different methods to achieve this (based on decompositions of orthogonal matrices, Givens decompositions, or Householder decompositions). Alternatively, when interpolating between planes the rotation is constructed towards any basis that represents the plane corresponding to the target frame, such that it follows the shortest path, i.e. the geodesic, avoiding any in-plane rotation. The geodesic interpolation is derived using singular value decomposition and is implemented in the `tourr` package. [EW: this paragraph perhaps needs some rework.]

@@

For any method we need to ensure that each projection along the interpolated path is still orthonormal. As explained above this means a tour path is a trace along a Stiefel manifold. In the case of one-dimensional projections, the set of bases traces a path along the surface of a hypersphere of radius one, as illustrated in Figure \ref{fig:notation-target-gen}, and for two-dimensional projections projections fall onto a torus shape. We can use a tour to visualize such a path, as shown in Figure \ref{fig:interpolation-2d}.

@@

Transformations of the data are common in the preprocessing [EW: "preprocessing" means b/f projecting?], such as scaling, sphering, or a logarithmic transformation for skewed distributions. However, we may also want to transform the data after projecting onto lower dimensions, to correct for unwanted effects of the projections. One example is the piling effect that is observed when projecting a high-dimensional distribution [@Diaconis1984-mv]: projected points are approximately Gaussian in most views, and increasingly concentrating near the center as dimensionality increases. The sage display [@Laa2020-uv] proposes a solution to this issue, via a radial transformation of the projected data. This non-linear transformation is defined in each point of the projection plane and is sensitive to the overall scale of the data as well as the original dimensionality $p$. In addition, tuning parameters can be used to obtain a more (or less) aggressive redistribution of the points. This is illustrated in Figure \ref{fig:pollen} showing the application of the sage transformation to the classical pollen dataset [@Coleman1986-so]. The projection without any rescaling looks similar to what is found when rescaling with default options (left), and we can use either of the two tuning parameters to better resolve the distribution near the center and reveal the hidden structure (middle and right).

@@

This review highlights the history of the grand tour and its modern extensions for the visualization of high dimensional numerical data. A rich ecosystem of tour methods are being actively developed and are finding applications in a broad range of areas. Tours are being combined with interactive graphics to assist with cluster analysis (the spinifex and liminal packages), displays of projected data are augmented to allow the identification of local structures (the sage algorithm, slicing and section pursuit), and  new visualizations based on tour methods are used to better diagnose optimization procedures for projection pursuit (ferrn).

@@

* highlighting the recent developments, spinifex, liminal, sage, slicing and sectioning, ferrn etc. 
* what's next - conditional model visualisation (Catherine Hurley's work), a interactive graphics system that easily integrates tours that provides the direct manipulations we want, pre-processing / transformations for high dimensional data, penalised projection pursuit, different kinds of data and spaces

>>>>

The focus of this review will be on visualizing high-dimensional numerical data using linear projections, in particular, as provided by the grand tour [@Asimov1985-xr; @Buja2005-cx]. The reason being that it is not feasible to adequately review the very large area of visualizing high-dimensions, and there have been numerous developments in tours recently. An overview of the technique, and new modifications is provided, along with how these techniques can be used in a variety of applications.

While tours are invaluable for assessing the geometry of data, they are by no means the only technique available for visualizing structure in high dimensional data. An early technique proposed for assessing pairwise relationships between variables is the scatterplot matrix (SPLOM) [@Tukey1981-ov; @Chambers1983-dh; @Tukey1983-fj; @Carr1984-xw; @Becker1987-dw]. The SPLOM allows the viewer to assess correlation structure but does not scale to large numbers of variables. Similarly, parallel coordinates plots (PCP) can be used to explore correlation and collinearity [@Inselberg1985-kf; @Wegman1990-dy]. By placing multiple variables side by side in a PCP higher order structure like clustering or lower dimensional embeddings are revealed, however the ordering of variables along the axis changes what can be learned. Another display that relies on variable ordering is the heatmap, which is widely used to visualize cluster structure in bioinformatics. @Wilkinson2009-pj provides a comprehensive history of this display in the social and natural sciences.

All of the aforementioned techniques can be enhanced through the use of interactivity.
By combining views with interactive elements like tool tips or highlighting the analysts is able to quickly interrogate interesting features of the data. One particularly important interaction technique in the history of statistical graphics is called brushing [@Becker1987-dw]. When brushing an analyst drags their mouse over the view which results in a region being drawn onto the canvas. When there are multiple views present the act of brushing can be thought of as a database query; points that fall inside the brush can be used to highlight or filter data on adjacent views (Figure \ref{fig:brushing}). This technique is particularly useful when combined with the tour (Section \@ref(sec:spin)).

@@

When using a tour, a sequence of $d$-dimensional linear projections are obtained from a $p$-dimensional space, where $d \ll p$. Let $X_{n \times p}$ be the data matrix, consisting of $n$ observations and $p$ variables, whose projections are of interest. A projection basis, $A_{p \times d}$, is matrix that characterizes the direction from which the data are projected and needs to satisfy an orthonormality condition, which requires that each column in $A$ has unit length and are perpendicular. <!-- This condition gives the $p$-dimensional basis space a geometrical structure of a Stiefel manifold. (Not sure this adds anything) --> With a data matrix and a projection basis, a projection of the data can be defined as $Y = X \cdot A$. Figure \ref{fig:notation-huber} shows an example of lower dimensional projections of the palmer penguins data in a Huber plot and a histogram [@penguins].

<!-- SL: I think there needs to be more detail here to link to the next paragraph and more explanation about what the huber plot is showing -->

\begin{figure*}
   \centering
         \includegraphics[width=\textwidth]{figures/notation-huber.pdf}
\caption{Huber's plot (left) of the penguins data (bill length and depth only). The solid ellipse represents how well the projection, onto different directions, separates the three species and the dashed circle is a reference guide. The histogram (right) is the 1D projection of the data onto the direction outlined as the solid line in the huber's plot. Adapted from E.-K. Lee et al. (2005), Figure 1. }
\label{fig:notation-huber}
\end{figure*}

@@

<!-- SL: I think this paragraph could moved down to the bottom of this section -->
Various displays are available to show projections in 1-$d$, 2-$d$ and higher dimensions. A 1-$d$ projection displays the data in a histogram analogous to a shadow puppet being projected onto a wall, while 2-$d$ projections are displayed as scatterplots. Higher dimensional projections can be shown with multivariate displays like SPLOMs or parallel coordinates. Figure \ref{fig:display} shows a variety of displays for projections and these displays are useful to show the non-normal distributions in the projected data, clustering structure, and multivariate relationship between variables.

@@

<!-- SL: I think this needs to be linked up to the notation section -->

@@

Instead of randomly generating targets, guided tour [@Cook1995-ae] finds its targets with a technique called projection pursuit [@Friedman1974-ck].  In projection pursuit, an index function is defined that is a surjective mapping from the projection $Y$ in $\mathbb{R}^{n \times d}$ to a scalar in $\mathbb{R}$. The guided tour iteratively optimizes the index function in the projection pursuit and produces the most interesting projection according to the index measure. The right panel of Figure \ref{fig:notation-target-gen} shows an example of a guided tour path.

@@

The smooth animation between the selected target frames is generated via a step-wise rotation in the high-dimensional space. There are two different methods to derive this rotation, first described in @Buja2005-cx., When interpolating between frames we rotate the starting basis onto the exact basis representation selected as a basis, and @Buja2005-cx describes three different methods to achieve this (based on decompositions of orthogonal matrices, Givens decompositions, or Householder decompositions) <!--SL : this sentence is confusing -->. Alternatively, when interpolating between planes the rotation is constructed towards any basis that represents the plane corresponding to the target frame, such that it follows the shortest path, i.e. the geodesic, avoiding any in-plane rotation. The geodesic interpolation is derived using singular value decomposition and is implemented in the `tourr` package.

<!-- SL: The above paragraph is very confusing to me, what is a plane and frame? what is a geodesic? why should it be a geodesic? -->

@@

For any method we need to ensure that each projection along the interpolated path is still orthonormal. As explained above this means a tour path is a trace along a Stiefel manifold. <!-- I still don't know if this adds anything, without it being defined --> In the case of one dimensional projections, the set of bases traces a path along the surface of a hypersphere of radius one, as illustrated in Figure \ref{fig:notation-target-gen}, and for two dimensional projections projections fall onto a torus shape. We can use a tour to visualize such a path, as shown in Figure \ref{fig:interpolation-2d}.

@@

Transformations of data are often useful prior to touring such as scaling, sphering, or a logarithmic transformation for skewed distributions. However, we may also want to transform the data after projecting onto lower dimensions, to correct for unwanted effects of the projections. One example is the piling effect that is observed when projecting a high-dimensional distribution [@Diaconis1984-mv]: projected points are approximately Gaussian in most views, and increasingly concentrating near the center as dimensionality increases. The sage display [@Laa2020-uv] proposes a solution to this issue, via a radial transformation of the projected data. This nonlinear transformation is defined in each point of the projection plane and is sensitive to the overall scale of the data as well as the original dimensionality $p$. In addition, tuning parameters can be used to obtain a more (or less) aggressive redistribution of the points. This is illustrated in Figure \ref{fig:pollen} showing the application of the sage transformation to the classical pollen dataset [@Coleman1986-so]. The projection without any rescaling looks similar to what is found when rescaling with default options (left), and we can use either of the two tuning parameters to better resolve the distribution near the center and reveal the hidden structure (middle and right).

@@

This review highlights the history of the grand tour and its modern extensions for the visualization of high dimensional numerical data. A rich ecosystem of tour methods are being actively developed and are finding applications in a broad range of areas such as machine learning and physics. Tours are being combined with interactive graphics to assist with cluster analysis (the spinifex and liminal packages), displays of projected data are augmented to allow the identification of local structures (the sage algorithm, slicing and section pursuit), and  new visualizations based on tour methods are used to better diagnose optimization procedures for projection pursuit (ferrn).

@@

<!--
* highlighting the recent developments, spinifex, liminal, sage, slicing and sectioning, ferrn etc. 
* what's next - conditional model visualisation (Catherine Hurley's work), a interactive graphics system that easily integrates tours that provides the direct manipulations we want, pre-processing / transformations for high dimensional data, penalised projection pursuit, different kinds of data and spaces
-->
